Optimal updating factor in adaptive flat-distribution-sampling simulations

### 2017-04-14 doc/invt.tex
add the referee reports

We will mainly focus on the adaptive methods below.
%
In the WL algorithm\cite{wang2001, *wang2001pre},
the bias potential, $V_i(t)$, is updated
at each sampling step $t$ as
%
\begin{equation}
  V_i(t+1)
  =
  V_i(t)
  +
  \delta_{i, \, i(t)}
  \frac{ \alpha(t) } { p_i }
  ,
\label{eq:wl_update}
\end{equation}
%
where $i(t)$ is the bin at step $t$,
$\delta_{i, \, i(t)}$ is the Kronecker delta function,
which is $1$ when $i = i(t)$ or $0$ otherwise,
and $\alpha(t)$ is the updating magnitude.
%
We refer to this scheme as the \emph{single-bin scheme},
for it applies only to the bias potential
at the current bin, $i(t)$.
%
In a more general \emph{multiple-bin scheme}, however,
several neighboring bins are updated:
%
\begin{equation}
  V_i(t+1)
  =
  V_i(t)
  +
  w_{i, \, i(t)}
  \frac{ \alpha(t) }
       { p_{i(t)} },
  \label{eq:mbin_update}
\end{equation}
%
where the elements, $w_{ij}$'s, are a set of coefficients satisfying
the restrictions to be discussed
in Sec. \ref{sec:updating-matrix}.
%
The Gaussian updating scheme
used in metadynamics
is a multiple-bin scheme.

@@

By the translational invariance,
homogeneous updating schemes
satisfy detailed balance,
Eq. \eqref{eq:w_detailedbalance},
for the flat distribution, $\rho_i = 1/n$,
i.e. the updating matrix, $\mathbf w$,
is symmetric.
%
The updating matrices
share the same set of eigenvectors
(which here are sines and cosines of different frequencies),
and thus are completely determined
by the eigenvalues.
%
Further, the multiplication of $\mathbf w$
can be reduced to a convolution with the window function,
or the updating kernel.
%
Thus,
there is a one-to-one correspondence between
the eigenvalues of $\mathbf w$
and the updating kernel.

@@

It is natural to first introduce homogeneous updating schemes
for a periodic variable\cite{dama2014}.
%
The updating matrix, $\mathbf w$,
assumes the following form
%
\begin{equation}
  w_{ij}
  =
  \mu_{i-j}
  +
  \mu_{i-j+n}
  +
  \mu_{i-j-n}
  ,
\notag
%\label{eq:w_band_pbc}
\end{equation}
%
where
$\mu_{-b}, \dots, \mu_b$ ($b \le n/2$)
are a set of numbers satisfying
%
\begin{equation}
  \mu_{-b} + \cdots + \mu_b = 1
  ,
\label{eq:mu_normalization}
\end{equation}
%
and $\mu_l = 0$ for $|l| > b$.
%
We will further impose the symmetry
%
\begin{equation}
  \mu_i = \mu_{-i}
  .
\label{eq:mu_symm}
\end{equation}
%
If $b = n/2$, then $\mu_{-b}$ is removed
from the sum in Eq. \eqref{eq:mu_normalization}
to avoid double counting.
%
These numbers constitute an \emph{updating kernel}\cite{bussi2006}.
%
In this way, the matrix, $\mathbf w$,
being symmetric with respect to the indices, $i$ and $j$,
satisfies Eqs. \eqref{eq:w_sumj}-\eqref{eq:w_balance},
for the flat distribution.

@@

Parallel to Eq. \eqref{eq:phi_pbc},
we define the out-of-boundary values
of $\phi_i$ as
%
\begin{equation}
  \phi_i
  =
  \begin{dcases}
    \phi_{ 1 - i }           & \mathrm{for \;} i \le 0, \\
    \phi_{ 2 \, n + 1 - i }  & \mathrm{for \;} i > n,
  \end{dcases}
\label{eq:phi_refl}
\end{equation}
%
such that the matrix multiplication is still equivalent to
a convolution as Eq. \eqref{eq:wmul_to_convol}.
%
\note{The derivation of Eq. \eqref{eq:wmul_to_convol}
  in this case is similar,
  $$
  \begin{aligned}
    \sum_{j = 1}^n w_{ij} \, \phi_j
    &=
    \sum_{j = 1}^n
      \mu_{i - j} \, \phi_j
    +
    \sum_{j = 1}^n
      \mu_{i + j - 1} \, \phi_j
    +
    \sum_{j = 1}^n
      \mu_{i + j - 2 \, n - 1} \, \phi_j
    \\
    &=
    \sum_{j = 1}^n
      \mu_{i - j} \, \phi_j
    +
    \sum_{l = 1 - n}^0
      \mu_{i - l} \, \phi_l
    +
    \sum_{l = n + 1}^{ 2 \, n }
      \mu_{i - l} \, \phi_l
    \\
    &=
    \sum_{j = 1 - n}^{ 2 \, n}
      \mu_{i - j} \, \phi_j.
  \end{aligned}
  $$
}%
We note that Eq. \eqref{eq:phi_refl}
defines a periodic variable of period $2 \, n$
with a reflective symmetry around $i = 1/2$.
%
Accordingly, the eigenvectors,
$\pmb\phi^{(1)}, \dots, \pmb\phi^{(n)}$,
must share the same periodicity,
and be even functions about the axis, $i = 1/2$:
%
\begin{equation}
  \phi^{(k)}_i
  =
  \phi_{k i}
  =
  \frac{ \sqrt{ 2 - \delta_{k, 1} } }
       {             n              }
  \cos \frac{ ( k - 1) \, \left( i - \frac 1 2 \right) \, 2 \, \pi}
            {                    2 \, n                           }
  ,
\notag
%\label{eq:wband_eigenvector_refl}
\end{equation}
%
The eigenvalues can be borrowed from
Eq. \eqref{eq:wband_eigenvalue_pbc},
with the period, $n$, doubled:
%
\begin{align}
  \lambda_k
  &=
  \mu_0
  +
  2
  \sum_{l = 1}^b
    \mu_l
    \cos \frac{(k - 1)  \, l \, 2 \, \pi}{ 2 \, n }
  .
\label{eq:wband_eigenvalue_refl}
\end{align}
%
\note{Derivation.
  First consider the unnormalized eigenvectors,
  $$
  \Phi^{(k)}_i
  =
  \cos \frac{ \left( i - \frac 1 2 \right) (k - 1) \, \pi}{n},
  $$
  which satisfies Eq. \eqref{eq:phi_refl}.
  %
  So, by Eq. \eqref{eq:wmul_to_convol},
  $$
  \begin{aligned}
  \sum_{i = 1}^n
    \Phi^{(k)}_i \, w_{ij}
  &=
  \sum_{l = -b}^b
    \Phi^{(k)}_{j - l} \, \mu_l
  \\
  &=
    \Phi^{(k)}_j \, \mu_0
  + \sum_{l=0}^{b}
    \mu_l \,
    \left[
      \Phi^{(k)}_{j-l}
      +
      \Phi^{(k)}_{j+l}
    \right]
  \\
  &= \Phi^{(k)}_j \, \lambda_k,
  \end{aligned}
  $$
  with $\lambda_k$ given by Eq. \eqref{eq:wband_eigenvalue_refl}.
}%
%
We may unify Eqs. \eqref{eq:wband_eigenvalue_refl}
and \eqref{eq:wband_eigenvalue_pbc} as
%
\begin{equation}
  \lambda_k
  =
  \mu_0
  +
  2
  \sum_{ l = 1 }^b
    \mu_l \,
    \cos \frac{ (k - 1) \, l \, g \, \pi }
              {            n             }
  ,
  \label{eq:wband_eigenvalue}
\end{equation}
%
where the degeneracy, $g$, is $2$
for the periodic case,
or $1$ otherwise.
%
In both cases,
the eigenvalues, $\lambda_k$'s,
are the cosine transform of
the updating kernel.
%
For the updating scheme to be stable,
the cosine transform of the kernel
must be nonnegative.
%
If this is not so,
we can modify the kernel
using the technique described
in Appendix \ref{sec:stabilize_wband}.

@@

\section{\label{sec:stabilize_wband}
Stabilization of homogeneous updating schemes}

>>>>

We will mainly focus on the adaptive methods below.
%
In the WL algorithm\cite{wang2001, *wang2001pre},
the bias potential, $V_i(t)$, is updated
at each sampling step $t$ as
%
\begin{equation}
  V_i(t+1)
  =
  V_i(t)
  +
  \delta_{i, \, i(t)}
  \frac{ \alpha(t) } { p_{i(t)} }
  ,
\label{eq:wl_update}
\end{equation}
%
where $i(t)$ is the bin at step $t$,
$\delta_{i, \, i(t)}$ is the Kronecker delta function,
which is $1$ when $i = i(t)$ or $0$ otherwise,
and $\alpha(t)$ is the updating magnitude.
%
We refer to this scheme as the \emph{single-bin scheme},
for it applies only to the bias potential
at the current bin, $i(t)$.
%
In a more general \emph{multiple-bin scheme}, however,
several neighboring bins are updated:
%
\begin{equation}
  V_i(t+1)
  =
  V_i(t)
  +
  w_{i, \, i(t)}
  \frac{ \alpha(t) }
       { p_{i(t)} },
  \label{eq:mbin_update}
\end{equation}
%
where the elements, $w_{ij}$'s, are a set of coefficients satisfying
the restrictions to be discussed
in Sec. \ref{sec:updating-matrix}.
%
The Gaussian updating scheme
used in metadynamics
is a multiple-bin scheme.

@@

@@

%It is natural to first introduce homogeneous updating schemes
For a periodic variable\cite{dama2014},
the updating matrix, $\mathbf w$,
assumes the following form
%
\begin{equation}
  w_{ij}
  =
  \mu_{i-j}
  +
  \mu_{i-j+n}
  +
  \mu_{i-j-n}
  ,
\notag
%\label{eq:w_band_pbc}
\end{equation}
%
where
$\mu_{-b}, \dots, \mu_b$ ($b \le n/2$)
are a set of numbers satisfying
%
\begin{equation}
  \mu_{-b} + \cdots + \mu_b = 1
  ,
\label{eq:mu_normalization}
\end{equation}
%
and $\mu_l = 0$ for $|l| > b$.
%
We will further impose the symmetry
%
\begin{equation}
  \mu_i = \mu_{-i}
  .
\label{eq:mu_symm}
\end{equation}
%
If $b = n/2$, then $\mu_{-b}$ is removed
from the sum in Eq. \eqref{eq:mu_normalization}
to avoid double counting.
%
These numbers constitute an \emph{updating kernel}\cite{bussi2006}.
%
In this way, the matrix, $\mathbf w$,
being symmetric with respect to the indices, $i$ and $j$,
satisfies Eqs. \eqref{eq:w_sumj}-\eqref{eq:w_balance},
for the flat distribution.

@@

Parallel to Eq. \eqref{eq:phi_pbc},
we define the out-of-boundary values
of $\phi_i$ as
%
\begin{equation}
  \phi_i
  =
  \begin{dcases}
    \phi_{ 1 - i }           & \mathrm{for \;} i \le 0, \\
    \phi_{ 2 \, n + 1 - i }  & \mathrm{for \;} i > n,
  \end{dcases}
\label{eq:phi_refl}
\end{equation}
%
such that the matrix multiplication is still equivalent to
a convolution as Eq. \eqref{eq:wmul_to_convol}.
%
\note{The derivation of Eq. \eqref{eq:wmul_to_convol}
  in this case is similar,
  $$
  \begin{aligned}
    \sum_{j = 1}^n w_{ij} \, \phi_j
    &=
    \sum_{j = 1}^n
      \mu_{i - j} \, \phi_j
    +
    \sum_{j = 1}^n
      \mu_{i + j - 1} \, \phi_j
    +
    \sum_{j = 1}^n
      \mu_{i + j - 2 \, n - 1} \, \phi_j
    \\
    &=
    \sum_{j = 1}^n
      \mu_{i - j} \, \phi_j
    +
    \sum_{l = 1 - n}^0
      \mu_{i - l} \, \phi_l
    +
    \sum_{l = n + 1}^{ 2 \, n }
      \mu_{i - l} \, \phi_l
    \\
    &=
    \sum_{j = 1 - n}^{ 2 \, n}
      \mu_{i - j} \, \phi_j.
  \end{aligned}
  $$
}%
We note that Eq. \eqref{eq:phi_refl}
defines a periodic variable of period $2 \, n$
with a reflective symmetry around $i = 1/2$.
%
Accordingly, the eigenvectors,
$\pmb\phi^{(1)}, \dots, \pmb\phi^{(n)}$,
must share the same periodicity,
and be even functions about the axis, $i = 1/2$:
%
\begin{equation}
  \phi^{(k)}_i
  =
  \phi_{k i}
  =
  \frac{ \sqrt{ 2 - \delta_{k, 1} } }
       {             n              }
  \cos \frac{ ( k - 1) \, \left( i - \frac 1 2 \right) \, 2 \, \pi}
            {                    2 \, n                           }
  ,
\notag
%\label{eq:wband_eigenvector_refl}
\end{equation}
%
The eigenvalues can be borrowed from
Eq. \eqref{eq:wband_eigenvalue_pbc},
with the period, $n$, doubled:
%
\begin{align}
  \lambda_k
  &=
  \mu_0
  +
  2
  \sum_{l = 1}^b
    \mu_l
    \cos \frac{(k - 1)  \, l \, 2 \, \pi}{ 2 \, n }
  .
\label{eq:wband_eigenvalue_refl}
\end{align}
%
\note{Derivation.
  First consider the unnormalized eigenvectors,
  $$
  \Phi^{(k)}_i
  =
  \cos \frac{ \left( i - \frac 1 2 \right) (k - 1) \, \pi}{n},
  $$
  which satisfies Eq. \eqref{eq:phi_refl}.
  %
  So, by Eq. \eqref{eq:wmul_to_convol},
  $$
  \begin{aligned}
  \sum_{i = 1}^n
    \Phi^{(k)}_i \, w_{ij}
  &=
  \sum_{l = -b}^b
    \Phi^{(k)}_{j - l} \, \mu_l
  \\
  &=
    \Phi^{(k)}_j \, \mu_0
  + \sum_{l=0}^{b}
    \mu_l \,
    \left[
      \Phi^{(k)}_{j-l}
      +
      \Phi^{(k)}_{j+l}
    \right]
  \\
  &= \Phi^{(k)}_j \, \lambda_k,
  \end{aligned}
  $$
  with $\lambda_k$ given by Eq. \eqref{eq:wband_eigenvalue_refl}.
}%
%
We may unify Eqs. \eqref{eq:wband_eigenvalue_refl}
and \eqref{eq:wband_eigenvalue_pbc} as
%
\begin{equation}
  \lambda_k
  =
  \mu_0
  +
  2
  \sum_{ l = 1 }^b
    \mu_l \,
    \cos \frac{ (k - 1) \, l \, g \, \pi }
              {            n             }
  ,
  \label{eq:wband_eigenvalue}
\end{equation}
%
where the degeneracy, $g$, is $2$
for the periodic case,
or $1$ otherwise.
%
In both cases,
the eigenvalues, $\lambda_k$'s,
are the cosine transform of
the updating kernel.

For the updating scheme to be stable,
the cosine transform of the kernel
must be nonnegative.
%
If this is not so,
we can modify the kernel
using the technique described
in Appendix \ref{sec:stabilize_wband}.

@@

\subsection{\label{sec:more_wband}
Homogeneous updating schemes
}

Here, we present some detailed derivations
on the homogeneous updating schemes.
%
By the translational invariance,
homogeneous updating schemes
satisfy detailed balance,
Eq. \eqref{eq:w_detailedbalance},
for the flat distribution, $\rho_i = 1/n$,
i.e. the updating matrix, $\mathbf w$,
is symmetric.
%
The updating matrices
share the same set of eigenvectors
(which here are sines and cosines of different frequencies),
and thus are completely determined
by the eigenvalues.
%
Further, the multiplication of $\mathbf w$
can be reduced to a convolution with the window function,
or the updating kernel.
%
Thus,
there is a one-to-one correspondence between
the eigenvalues of $\mathbf w$
and the updating kernel.

\subsection{\label{sec:stabilize_wband}
Stabilization}
